---
title: "Chapter5_AdditionalTimeDomainTopics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## load required packages
```{r}
library('astsa')
source('grid.r')
```
### Example 5.1 Long Memory Fitting of the Glacial Varve Series
```{r}
ACF  =  acf(log(varve), 100, plot=FALSE)
plot(ACF[1:100], xlab="LAG", ylim=c(-.05,1), panel.first=grid(lty=1))
abline(h=0)
```
```{r}
d  = 0.3841688
p = rep(1,31)
for (k in 1:30){ p[k+1] = (k-d)*p[k]/(k+1) }
plot(1:30, p[-1], ylab=expression(pi(d)), lwd=2, xlab="Index", type="h", panel.first=grid(lty=1))
```
```{r message=FALSE}
library(fracdiff)
# I might suggest another package such as 'arfima' because 
# this package gives questionable results and it's not easy to
# pull out the residuals
lvarve = log(varve)-mean(log(varve))
varve.fd = fracdiff(lvarve, nar=0, nma=0, M=30)
#varve.fd$d  # = 0.3841688
#varve.fd$stderror.dpq  # = 4.589514e-06 (questionable result!!)
res.fd = diffseries(log(varve), varve.fd$d)  # frac diff resids
res.arima = resid(arima(log(varve), order=c(1,1,1))) # arima resids
par(mfrow=c(2,1), mar=c(2,2.5,.5,0)+.5, mgp=c(1.4,.6,0))
acf(res.arima, 100, xlim=c(4,97), ylim=c(-.2,.2), main="", xlab="LAG", panel.first=grid(lty=1))
acf(res.fd, 100, xlim=c(4,97), ylim=c(-.2,.2), main="", xlab="LAG", panel.first=grid(lty=1))
```
```{r}
library(arfima)
summary(varve.fd <- arfima(log(varve)))  # d.hat = 0.3728, se(d,hat) = 0.0273
# residual stuff
innov = resid(varve.fd)  
plot.ts(innov[[1]])  
acf(innov[[1]])  

## ... much better ...  sorry I didn't ...
## ... get it in for the newest edition ..
## ... once in awhile, they slip on by ...
```
### Example 5.2 Long Memory Spectra for the Varve Series
```{r}
series = log(varve) - mean(log(varve)) # specify series to be analyzed
d0 = .1                                # initial value of d
n.per = nextn(length(series))
m=(n.per)/2  - 1
per = abs(fft(series)[-1])^2  # remove 0 freq
per = per/n.per               # R doesn't scale fft by sqrt(n)
g = 4*(sin(pi*((1:m)/n.per))^2)

# Function to calculate -log.likelihood
whit.like = function(d){
	g.d=g^d
    sig2 = (sum(g.d*per[1:m])/m)
    log.like = m*log(sig2) + d*sum(log(g)) + m
    return(log.like)
    }	

# -- Estimation --
est=optim(d0,whit.like,gr=NULL,method="L-BFGS-B",hessian=TRUE,
           lower=-.5,upper=.5)
#cat("d.hat =", est$par,"se(dhat) =", 1/sqrt(est$hessian), "\n")
g.dhat=g^est$par;  sig2=sum(g.dhat*per[1:m])/m
#cat("sig2hat=",sig2, "\n")  

#  spectral approach
u = spec.ar(log(varve), plot=FALSE)  #produces ar(8)
g= 4*(sin(pi*((1:500)/2000))^2)
fhat = sig2*g^{-est$par} # long memory spectral estimate

plot(1:500/2000, log(fhat), type="l", ylab="log(spectrum)", xlab="frequency",  panel.first=grid(lty=1))

lines(u$freq[1:250], log(u$spec[1:250]), lty="dashed")
```
```{r}
library(fracdiff)
fdGPH(log(varve), bandw = .9)  # fdGPH doesn't seem to work very well either
```
```{r}
# page - 249
ar.mle(log(varve)) # to get AR(8) estimates
```
### Example 5.3 Testing Unit Roots in the Glacial Varve Series
https://www.stat.pitt.edu/stoffer/tsa4/Rexamples.htm
```{r message=FALSE}
library(tseries)
adf.test(log(varve), k=0)  # DF test
adf.test(log(varve))       # ADF test
pp.test(log(varve))        # PP test
```
* In each test, we reject the null hypothesis that the logged varve series has a unit
root. The conclusion of these tests supports the conclusion of the previous section
that the logged varve series is long memory rather than integrated.

### Example 5.4 Analysis of U.S. GNP
```{r}
gnpgr = diff(log(gnp))          # get the returns
u     = sarima(gnpgr, 1, 0, 0)  # fit an AR(1)
```
```{r}
acf2(resid(u$fit), 20)          # get (p)acf of the squared residuals
```

```{r}
# {innov2acf}
x = resid(arima(diff(log(gnp)), order=c(1,0,0)))
ACF = acf(x^2, 20, plot=FALSE)$acf[-1]
PACF = pacf(x^2, 20, plot=FALSE)$acf
num = length(gnp)-1
minA = min(ACF)
maxA = max(ACF)
minP = min(PACF)
maxP = max(PACF)
U = 2/sqrt(num)
L = -U
LAG = 1:20/4
minu = min(minA, minP, L) - 0.01
maxu = min(maxA + 0.2, maxP + 0.2, 1)

par(mfrow=c(2,1), mar=c(2,2.5,0,0)+.5, mgp=c(1.4,.6,0))
plot(LAG, ACF, type="h", xlab="LAG", ylim = c(minu, maxu), panel.first=grid(lty=1)); abline(h=0)
abline(h = c(L, U), col=4, lty=2)  
plot(LAG, PACF, type="h", xlab="LAG",  ylim = c(minu, maxu) , panel.first=grid(lty=1)); abline(h=0)
abline(h = c(L, U), col=4, lty=2)  
```

```{r message=FALSE}
library(fGarch)
summary(garchFit(~arma(1,0)+garch(1,0), gnpgr))
```
### Example 5.5 GARCH Analysis of the DJIA Returns
```{r}
library(xts)   # needed to handle djia
djiar = diff(log(djia$Close))[-1]
7
par(mfrow=c(2,1))
acf2(djiar)    # exhibits some autocorrelation (not shown)
acf2(djiar^2)  # oozes autocorrelation (not shown)
```
### Example 5.6 APARCH Analysis of the DJIA Returns
### not converge ???
```{r}
library(fGarch)
# GARCH fit
djia.g <- garchFit(~arma(1,0)+garch(1,1), data=djiar, cond.dist='std')
plot(djia.g)    # to see all plot options
```
```{r}
# APARCH fit
djia.ap <- garchFit(~arma(1,0)+aparch(1,1), data=djiar, cond.dist='std')
plot(djia.ap)
```
* github version
```{r}
 # djia = getYahooData("^DJI",start=20060420,end=20160420,freq="daily") 
library(fGarch)
summary(djia.g <- garchFit(~arma(1,0)+garch(2,1), data=djiar, cond.dist='std'))
u = djia.g@sigma.t

par(mar=c(2.5,2,0,.5)+.5, mgp=c(1.6,.6,0), oma=rep(0,4))
plot(djiar[400:900], type='n', main='')
lines(djiar[400:900], col=gray(.6))
lines((u+djiar-djiar)[400:900], col=4)
```
### Example 5.7 Threshold Modeling of the Influenza Series
```{r}
tsplot(flu, type="c")
Months = c("J","F","M","A","M","J","J","A","S","O","N","D")
points(flu, pch=Months, cex=.8, font=2)
```
```{r}
dflu = diff(flu)
thrsh = .05 # threshold
Z = ts.intersect(dflu, lag(dflu,-1), lag(dflu,-2), lag(dflu,-3), lag(dflu,-4))
ind1 = ifelse(Z[,2] < thrsh, 1, NA)  # indicator < thrsh
ind2 = ifelse(Z[,2] < thrsh, NA, 1)  # indicator >= thrsh
X1 = Z[,1]*ind1
X2 = Z[,1]*ind2
summary(fit1<-lm(X1~Z[,2:5]))  # case 1
```
```{r}
summary(fit2<-lm(X2~Z[,2:5]))  # case 2
```
```{r}
D = cbind(rep(1, nrow(Z)), Z[,2:5])  # get predictions
b1 = fit1$coef
b2 = fit2$coef
p1 = D%*%b1
p2 = D%*%b2
prd = ifelse(Z[,2] < thrsh, p1, p2)

plot(dflu, ylim=c(-.5,.5), type='n')
grid(lty=1)
points(dflu, pch=3)
lines(prd)
prde1 = sqrt(sum(resid(fit1)^2)/df.residual(fit1))
prde2 = sqrt(sum(resid(fit2)^2)/df.residual(fit2))
prde = ifelse(Z[,2] < thrsh, prde1, prde2)
x = time(dflu)[-(1:4)]
xx = c(x, rev(x))
yy = c(prd - 2*prde, rev(prd + 2*prde))

polygon(xx, yy, border=8, col=rgb(.6,.6,.6,alpha=.25))
abline(h=.05, col=4, lty=6)
```
### Example 5.8 Relating the Prewhitened SOI to the Transformed Recruitment Series
```{r}
soi.d   = resid(lm(soi~time(soi), na.action=NULL)) # detrended SOI
acf2(soi.d)
```
```{r}
fit     = arima(soi.d, order=c(1,0,0))
ar1     = as.numeric(coef(fit)[1]) # = 0.5875
soi.pw  = resid(fit)
rec.fil = filter(rec, filter=c(1, -ar1), sides=1)
ccf2(soi.pw, rec.fil, na.action=na.omit)
```
### Example 5.9 Transfer Function Model for SOI and Recruitment
```{r}
fish  = ts.intersect(rec, RL1=lag(rec,-1), SL5=lag(soi.d,-5))
(u    = lm(fish[,1]~fish[,2:3], na.action=NULL))
acf2(resid(u)) # suggests ar1

```
```{r}
(arx  = sarima(fish[,1], 1, 0, 0, xreg=fish[,2:3])) # final model
pred  = rec + resid(arx$fit) # 1-step-ahead predictions
ts.plot(pred, rec, col=c('gray90',1), lwd=c(7,1))
```
### Example 5.10 Pollution, Weather, and Mortality
```{r}
library(vars)
x = cbind(cmort, tempr, part)
summary(VAR(x, p=1, type="both"))  # "both" fits constant + trend
```
```{r}
VARselect(x, lag.max=10, type="both")
```
