---
title: "Chapter1_CharacteristicsofTimeSeries"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r message=FALSE}
library('astsa')
library(tidyverse)
source('grid.r')
```
# Fig.1.1.Johnson&Johnsonquarterlyearningspershare,84quarters,1960-Ito1980-IV.
```{r}
plot(jj, ylab="Quarterly Earnings per Share",   type='n')
lines(jj, type="o")
grid(lty=1)
```
# Fig. 1.2. Left: Initial deposits /tsaEZ
Fig. 1.2. Left: Initial deposits of $100, $150, $200, then $75, in quarter 1, 2, 3, then 4, over 20 years, with an annual growth rate of 10%; xt = (1+.10)x_t???4. Right: Logs of the quarterly values; log(xt) = log(1+.10) +log(xt???4). 
```{r}
t=1:(10*2)
x = 100*1.1^t
y = 150*1.1^t
z = 200*1.1^t
w = 75*1.1^t 
culer <- c(1,2,'darkgreen',4)
u <- t(cbind(x,y,z,w))
x <- ts(c(u), freq=4)
```
```{r}
par(mfrow=1:2, mar=c(2.5,2.5,.5,.5)+.1, mgp=c(1.5,.6,0))
plot(x, xlab='quarter', ylab='value', col=gray(.6))
points(x,  pch=c('1','2','3','4'), col=culer, cex=.8)
plot(log(x), xlab='quarter', ylab='log(value)', col=gray(.6))
points(log(x),  pch=c('1','2','3','4'), col=culer, cex=.8)
```
# Fig.1.3.Yearlyaverageglobaltemperaturedeviations(1880-2015)in???C.
```{r}
par(mar=c(3,3,1,1), mgp=c(1.6,.6,0))
plot(globtemp, ylab="Global Temperature Deviations",    type='n')   
grid(lty=1)
lines(globtemp, type='o')
```
# Fig. 1.3. Speech recording of the syllable aaa···hhh sampled at 10,000 points per second with n = 1020points. Show a small .1 second (1000 point) sample of recorded speech 
```{r}
par(mar=c(3,3,1,1), mgp=c(1.6,.6,0))
plot(speech,   type='n', ylim=c(0,4100))
grid(lty=1)
lines(speech) 
```
# Fig.1.4.ThedailyreturnsoftheDowJonesIndustrialAverage(DJIA)fromApril20,2006to April20,2016.



