---
title: "tsaEZ notes"
author: "ETF Researcher"
date: "March 11, 2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
* tsaEZ? http://www.stat.pitt.edu/stoffer/tsa4/

```{r}
# load package
library(astsa)
```
# Time Series Characteristics

## Example 1.7 Moving Averages and Filtering
```{r}
w = rnorm(500, 0, 1)
v = filter(w, sides = 2, rep(1/3, 3))
par(mfrow = c(2,1))
tsplot(w, main = "white noise")
tsplot(v, ylim = c(-3, 3),main = "moving average")
```

## Example 1.8 Autoregressions
```{r}
w = rnorm(550, 0, 1)
x = filter(w, filter = c(1, -0.9), method = "recursive")[-(1:50)]
tsplot(x, main = "autoregression")
```

## Example 1.9 Random Walk with Drift
```{r}
set.seed(154)
w = rnorm(200)
x = cumsum(w)
wd = w + 0.2
xd = cumsum(wd)

tsplot(xd, ylim = c(-5, 55), main = "random walk", ylab = "")
abline(a = 0, b = 0.2, lty = 2)
lines(x, col = 4)
abline(h = 9, col = 4, lty = 2)
```

## Example 1.22 Prediction Using Cross-Correlation
```{r}
x = rnorm(100)
y = lag(x, -5) + rnorm(100)
ccf(y, x, ylab = 'CCovF', type = 'covariance')
```

# 2.2 Exploratory Data Analysis

## Example 2.5 Differencing Chicken Prices
```{r}
fit <- lm(chicken ~ time(chicken), na.action = NULL)
par(mfrow = c(2, 1))
tsplot(resid(fit), main = "detrend")
tsplot(diff(chicken), main = 'first difference')
```

```{r}
par(mfrow = c(3, 1))
acf1(chicken, 48, main = 'chicken')
acf1(resid(fit), 48, main = 'detrended')
acf1(diff(chicken), 48, main = 'first difference')
```

## Example 2.7 Paleoclimatic Glacial Varves (transformation)
```{r}
par(mfrow = c(2, 1))
tsplot(varve, main = "varve", ylab = "")
tsplot(log(varve), main = "log(varve)", ylab = "")
```

## Example 2.8 Scatterplot Matrices, SOI and Recruitment

```{r}
lag1.plot(soi, 12)
```
```{r}
lag2.plot(soi, rec, 8)
```

## Example 2.9 Regression with Lagged Variables (cont)
```{r}
dummy = ifelse(soi < 0, 0, 1)
fish = ts.intersect(rec, soiL6 = lag(soi, -6), dL6 = lag(dummy, -6), dframe = TRUE)
summary(fit <- lm(rec ~ soiL6*dL6, data = fish, na.action = NULL))
```
```{r}
attach(fish)
plot(soiL6, rec)
lines(lowess(soiL6, rec), col= 4, lwd=2)
points(soiL6, fitted(fit), pch="+", col= 2)
```
```{r}
tsplot(resid(fit))
acf(resid(fit))
```

## Example 2.10 Using Regression to Discover a Signal in Noise
```{r}
set.seed(90210) # so you can reproduce these results
x = 2*cos(2*pi*1:500/50 + .6*pi) + rnorm(500,0,5)
z1 = cos(2*pi*1:500/50)
z2 = sin(2*pi*1:500/50)
summary(fit <- lm(x~ 0 + z1 + z2)) # zero to exclude the intercept
```
```{r}
par(mfrow=c(2,1))
tsplot(x, margins=.25)
tsplot(x, col=8, margins=.25, ylab=expression(hat(x)))
lines(fitted(fit), col=2)

```

## 2.3 Smoothing Time Series

## Example 2.11 Moving Average Smoother
```{r}
wgts = c(.5, rep(1,11), .5)/12
soif = filter(soi, sides=2, filter=wgts)
tsplot(soi)
lines(soif, lwd=2, col=4)
```

## Example 2.12 Kernel Smoothing
```{r}
tsplot(soi)
lines(ksmooth(time(soi), soi, "normal", bandwidth=1), lwd=2, col=4)
SOI = ts(soi, freq=1); tsplot(SOI) # the time scale matters (not shown)
lines(ksmooth(time(SOI), SOI, "normal", bandwidth=12), lwd=2, col=4)

```

## Example 2.13 Lowess
```{r}
tsplot(soi)
lines(lowess(soi, f=.05), lwd=2, col=4) # El Nino cycle
lines(lowess(soi), lty=2, lwd=2, col=2) # trend (using default span)

```

## Example 2.15 Classical Structural Modeling
```{r}
x = window(hor, start=2002) # data set in astsa version 1.7.1+
plot(decompose(x)) # not shown
plot(stl(x, s.window='per')) # not shown
plot(stl(x, s.window=15))

```

# ARIMA Models

## Example 3.2 The Sample Path of an AR(1) Process
```{r}
par(mfrow=c(2,1))
tsplot(arima.sim(list(order=c(1,0,0), ar=.9), n=100), ylab="x",
main=(expression(AR(1)~~~phi==+.9)))
tsplot(arima.sim(list(order=c(1,0,0), ar=-.9), n=100), ylab="x",
main=(expression(AR(1)~~~phi==-.9)))

```

## Example 3.3 The MA(1) Process
```{r}
par(mfrow = c(2, 1))
tsplot(arima.sim(list(order=c(0, 0, 1), ma=0.9), n = 100), 
       ylab = "x",
       main = (expression(MA(1) ~~~ theta == +0.5)))

tsplot(arima.sim(list(order=c(0, 0, 1), ma=-0.9), n = 100), 
       ylab = "x",
       main = (expression(MA(1) ~~~ theta == -0.5)))

```

## 3.5 Forecasting

```{r}
sarima(rec, 2, 0, 0)
```
```{r}
sarima.for(rec, 24, 2, 0, 0)
```

## 3.6 Integrated Models

## Example 3.26 IMA(1, 1) and EWMA
```{r}
set.seed(666)
x = arima.sim(list(order = c(0, 1, 1), ma = -0.8), n = 100)
x.ima = HoltWinters(x, beta = FALSE, gamma = FALSE)
plot(x.ima, main = "EWMA")
```

## 3.7 Building ARIMA Models
## Example 3.27 Analysis of GNP Data
```{r}
tsplot(gnp)
```
```{r}
acf2(gnp, 50)
```
```{r}
gnpgr = diff(log(gnp))
tsplot(gnpgr)
```
```{r}
acf2(gnpgr, 24)
```

```{r}
# AR(1) model
sarima(gnpgr, 1, 0, 0)
```
```{r}
# MA(2) model
sarima(gnpgr, 0, 0, 2)
```

## Example 3.28 Diagnostics for GNP Growth Rate Example
## Example 3.31 Model Choice for the U.S. GNP Series

* Use AIC or BIC

## 3.8 Regression with Autocorrelated Errors
### Example 3.32 Mortality, Temperature and Pollution

```{r}
trend = time(cmort)
temp = tempr - mean(tempr)
temp2 = temp^2
fit <- lm(cmort ~ trend + temp + temp2 + part, na.action = NULL)
```
```{r message=FALSE, warning=FALSE}
# check the AR
acf2(resid(fit), 52)
```
```{r}
sarima(cmort, 2, 0, 0, xreg = cbind(trend, temp, temp2, part))
```

### Example 3.33 Regression with Lagged Variables (cont)
```{r}
dummy = ifelse(soi < 0, 0, 1)
fish = ts.intersect(rec, soiL6=lag(soi, -6), dL6 = lag(dummy, -6),
                    dframe = TRUE)

summary(fit <- lm(rec ~ soiL6*dL6, data = fish, na.action = NULL))
```

```{r}
attach(fish)
tsplot(resid(fit))
```

```{r}
acf2(resid(fit)) # suggest AR2
```
```{r}
intract = soiL6 * dL6
sarima(rec, 2, 0, 0, xreg = cbind(soiL6, dL6, intract))

```
```{r}
detach(fish)
```

## 3.9 Seasonal ARIMA Models
### Example 3.34 A Seasonal AR Series
```{r}
# three years data with phi = 0.9
set.seed(666)
phi = c(rep(0, 11), 0.9)
sAR = arima.sim(list(order=c(12, 0, 0), ar = phi), n = 37)
sAR = ts(sAR, frequency = 12)
```
```{r}
layout(matrix(c(1,2, 1,3), nc=2))
par(mar=c(3,3,2,1), mgp=c(1.6, 0.6, 0))
plot(sAR, axes = FALSE, main = 'seasonal AR(1)', xlab = "year", type = 'c')
Months = c("J","F","M","A","M","J","J","A","S","O","N","D")
points(sAR, pch=Months, cex=1.25, font=4, col=1:4)
axis(1, 1:4)
abline(v=1:4, lty=2, col=gray(.6))
axis(2)
box()
ACF = ARMAacf(ar=phi, ma=0, 100)
PACF = ARMAacf(ar=phi, ma=0, 100, pacf=TRUE)
plot(ACF, type="h", xlab="lag", ylim=c(-.1,1))
abline(h=0)
plot(PACF, type="h", xlab="lag", ylim=c(-.1,1))
abline(h=0)

```
### Example 3.35 A Mixed Seasonal Model

* ARMA(0, 1) × (1, 0)12 model
```{r}
phi = c(rep(0, 11), 0.8)
ACF = ARMAacf(ar = phi, ma=-0.5, 50)[-1]
PACF = ARMAacf(ar=phi, ma=-0.5, 50, pacf = TRUE)

par(mfrow = c(1, 2))
plot(ACF, type = "h", xlab = "lag", ylim = c(-0.4, 0.8))
abline(h = 0)

plot(PACF, type = "h", xlab = "lag", ylim = c(-0.4, 0.8))
abline(h = 0)

```
```{r}
# example for birth rate in US
par(mfrow = c(3, 1), mar=c(2,2,0,0) + 1, mgp = c(1.6, 0.6, 0))
tsplot(birth)
tsplot(diff(log(birth)))
acf1(diff(log(birth)), 61)
```
```{r}
# try decompose
plot(decompose(birth))
```

### Example 3.37 Air Passengers
```{r}
x = AirPassengers
lx = log(x) # stablize the variance
dlx = diff(lx) # remove trend
ddlx = diff(dlx, 12) # remove seasonal trend
plot.ts(cbind(x, lx, dlx, ddlx), yax.flip = TRUE, main = "")
```

```{r}
acf2(ddlx, 50)
```

* first try an ARIMA(1, 1, 1) × (0, 1, 1)12 on the logged data:
```{r}
sarima(lx, 1,1,1, 0,1,1,12)
# AR parameter is not significant,
```

```{r}
sarima(lx, 0, 1, 1, 0, 1, 1, 12)
```

```{r}
sarima(lx, 1, 1, 0, 0, 1, 1, 12)
```

* forecast with logged data in 12 months
```{r}
sarima.for(lx, 12, 0, 1, 1, 0, 1, 1, 12)
```


# Chapter 4 Spectral Analysis and Filtering

## 4.2 Periodicity and Cyclical Behavior
### Example 4.1 A Periodic Series
```{r}
library(astsa)
x1 = 2 * cos(2*pi*1:100*6/100) + 3*sin(2*pi*1:100*6/100)
x2 = 4*cos(2*pi*1:100*10/100) + 5*sin(2*pi*1:100*10/100)
x3 = 6*cos(2*pi*1:100*40/100) + 7*sin(2*pi*1:100*40/100)
x = x1 + x2 + x3
```
```{r}
par(mfrow = c(2, 2))
tsplot(x1, ylim=c(-10, 10), main=expression(omega==6/100~~~A^2==13))
tsplot(x2, ylim=c(-10,10), main=expression(omega==10/100~~~A^2==41))
tsplot(x3, ylim=c(-10,10), main=expression(omega==40/100~~~A^2==85))
tsplot(x, ylim=c(-16,16), main="sum")
```

### Example 4.2 Estimation and the Periodogram
```{r}
n = length(x)
P = (4/n) * Mod(fft(x) / sqrt(n))^2
Fr = 0:99 / 100
tsplot(Fr, P, type = 'o', xlab = "frequency", ylab = "scaled periodogram")
```
### Example 4.3 Star Magnitude
```{r}
n = length(star)
par(mfrow=c(2,1), mar=c(3,3,1,1), mgp=c(1.6,.6,0))
tsplot(star, ylab = "star magnitude", xlab = "day")
Per = Mod(fft(star - mean(star)))^2 / n
Freq = (1:n-1)/n
plot(Freq[1:50], Per[1:50], type = "h", lwd = 3,
     ylab = "Periodogram", xlab = "Frequency")

u = which.max(Per[1:50])
uu = which.max(Per[1:50][-u])
text(0.05, 7000, "24 day cycle")
text(0.027, 9000, "29 day cycle")

```

```{r}
# another way to find the two peaks is to order on Per
y = cbind(1:50, Freq[1:50], Per[1:50])
tail(y[order(y[, 3]), ], 10)
```
## 4.3 The Spectral Density
### Example 4.4 White Noise Series, Moving Average, AR(2)
```{r}
par(mfrow=c(3,1))
arma.spec(log="no", main="White Noise")
arma.spec(ma=0.5, log="no", main = "moving average")
arma.spec(ar=c(1.0, -0.9), log = "no", main = "autoregression")
```

## 4.4 Periodogram and Discrete Fourier Transform

```{r}
(dft = fft(1:4)/sqrt(4))
```
```{r}
(idft = fft(dft, inverse=TRUE)/sqrt(4))
```
```{r}
(Re(idft)) 
```
### Example 4.7 Periodogram of SOI and Recruitment Series
```{r}
par(mfrow=c(2,1)) # raw periodogram
mvspec(soi, log="no"); abline(v=1/4, lty="dotted")
mvspec(rec, log="no"); abline(v=1/4, lty="dotted")
```

## 4.5 Nonparametric Spectral Estimation
```{r}
library(astsa)
arma.spec(ar=c(1,-.9), xlim=c(.15,.151), n.freq=100000)
```

### Example 4.8 Averaged Periodogram for SOI and Recruitment
```{r}
par(mfrow=c(2,1))
(k = kernel("daniell", 4))
soi.ave = mvspec(soi, k, log="no"); abline(v=c(.25,1,2,3), lty=2)
rec.ave = mvspec(rec, k, log="no"); abline(v=c(.25,1,2,3), lty=2)

```

```{r}
soi.ave$bandwidth # = 0.225
soi.ave$df # = 16.9875
```
### Example 4.9 Harmonics
```{r}
t = seq(0, 1, by = 1/200)
amps = c(1, .5, .4, .3, .2, .1)
x = matrix(0, 201, 6)
for (j in 1:6) {
        x[,j] = amps[j]*sin(2*pi*t*2*j)
        }
x = ts(cbind(x, rowSums(x)), start=0, deltat=1/200)
```
```{r}
# no color ?
ts.plot(x, lty=c(1:6, 1), lwd=c(rep(1,6), 2), ylab="Sinusoids")
names = c("Fundamental","2nd Harmonic","3rd Harmonic","4th Harmonic", "5th
Harmonic", "6th Harmonic", "Formed Signal")
legend("topright", names, lty=c(1:6, 1), lwd=c(rep(1,6), 2) )
```

### Example 4.10 Smoothed Periodogram for SOI and Recruitment
```{r}
kernel("modified.daniell", c(3,3)) # for a list
plot(kernel("modified.daniell", c(3,3))) # for a plot
```
```{r}
k = kernel("modified.daniell", c(3,3))
soi.smo = mvspec(soi, k, taper=.1, log="no") # a taper is used
abline(v=c(1/4,1), lty="dotted")

```

### Example 4.11 The Effect of Tapering the SOI Series
```{r}
s0 = mvspec(soi, spans=c(7,7), plot=FALSE) # no taper
s50 = mvspec(soi, spans=c(7,7), taper=.5, plot=FALSE) # full taper
plot(s0$freq, s0$spec, log="y", type="l", lty=2, ylab="spectrum",
xlab="frequency") # dashed line
lines(s50$freq, s50$spec) # solid line
```


# Chapter 5 Some Additional Topics

## 5.1 GARCH Models

```{r}
# modeling on returns
tsplot(diff(log(gnp)), type = 'o')
points(diff(gnp)/lag(gnp, -1), pch = '+')
```

### Example 5.1 Analysis of U.S. GNP
```{r}
res = resid(sarima(diff(log(gnp)), 1, 0, 0, details = FALSE)$fit)
acf2(res^2, 20)
```

```{r}
library(fGarch)
gnpr = diff(log(gnp))
summary(garchFit(~ arma(1, 0) + garch(1, 0), data = gnpr))
```

### Example 5.2 GARCH Analysis of the DJIA Returns
```{r}
library(xts)
djiar = diff(log(djia$Close))[-1]
acf2(djiar)
```
```{r}
u = resid(sarima(djiar, 1, 0, 0, details = FALSE)$fit)
acf2(u^2)
```
```{r}
library(fGarch)
summary(djia.g <- garchFit(~arma(1, 0) + garch(1,1), data = djiar, cond.dist = 'std'))
```
```{r}
plot(djia.g, which = 3)
```

### Example 5.3 APARCH Analysis of the DJIA Returns
```{r message=FALSE, warning=FALSE}
library(fGarch)
library(xts)
```
```{r message=FALSE, warning=FALSE}
lapply(c('xts', 'fGarch'), library, char = TRUE)
```
```{r}
djiar = diff(log(djia$Close))[-1]

summary(djia.ap <- garchFit(~arma(1, 0) + aparch(1, 1), 
                        data = djiar, cond.dist = 'std'))
```
```{r}
plot(djia.ap)
```

## 5.2 Unit Root Testing

### Example 5.4 Testing Unit Roots in the Glacial Varve Series

```{r}
library(tseries)
```
```{r}
adf.test(log(varve), k=0)
```
```{r}
adf.test(log(varve))
```
```{r}
pp.test(log(varve))
```
## 5.3 Long Memory and Fractional Differencing

### Example 5.5 Long Memory Fitting of the Glacial Varve Series

* Compare residuals
```{r}
library(arfima)
```
```{r}
summary(varve.fd <- arfima(log(varve), order = c(0, 0, 0)))
```
```{r}
innov = resid(varve.fd)
tsplot(innov[[1]])
```

```{r}
acf1(innov[[1]])
```

```{r}
sarima(log(varve), 1, 1, 1)
```
